Quantifying and controlling biases in dark matter halo concentration estimates

### 2016-08-25 paper/ms.tex
new references and impact on recent work

The concentration of the NFW profile provides a conceptual framework
to study simulated dark matter halos as a function of redshift and
cosmological parameters.  In many numerical studies
\citep{Neto2007,Maccio2008,Duffy2008,Munoz2011,Prada2012,Ludlow2014}
the results are summarized through the mass-concentration
relationship; that is the distribution of concentration values at a
fixed halo mass and redshift.  The success of such numerical
experiments rests on a reliable algorithm to estimate the
concentration.  Such an algorithm should provide unbiased results and
must be robust when applied to simulations of different numerical
resolution.

@@

To date, there are two standard methods to estimate concentrations in
dark matter halos extracted from N-body simulations.  The first method
tries to directly estimate the density profile.  It takes all the
particles in the halo and bins them in the logarithm of the radial
coordinate from the halo center.  Then, it estimates the density in
each logarithmic bin counting the particles and dividing by the
corresponding shell volume.  At this point is possible to make a
direct fit to the density as a function of the radial coordinate.
This method has been broadly used for more than two decades to study
the mass-concentration-redshift relation of dark matter halos.

A second method uses the circular velocity profile.  It finds the
value of $x$ for which the normalized circular velocity $v(<x)$ shows
a maximum.  Using this value it solves numerically for the
corresponding value of the concentration using Eq. (\ref{eq:max_v}).
This method has been most recently used by \cite{Klypin2016} to study
the mass-concentration-redshift relation using the Multidark
Simulation Suite.

@@

In our bootstrapping experiments we take halos sampled with at least
$10^{5}$ particles and subsample each halo by factors of $2$ up to
$10^{3}$ to measure the concentration every time the halo is
resampled. 
We use a two-sample Kolmogorov-Smirnov (KS) test to check that the
bootstrapped sets are a fair subsample of the parent halo.
Using the KS test we compare the list of radial distances from each
subsample against that of its parent halo.
We find that the resulting p-value distribution is flat.
This confirms that the radial distribution of particles in the
bootraped halo is consistent with coming from the distribution given
by the parent halo.

@@

@@

 The authors acknowledge the technical support from the new
 high-performance computing facility at Uniandes. JEFR acknowledges
 financial support from Vicerrector\'ia de Investigaciones at Uniandes
 through a FAPA project. JCMC acknowledges financial support from
 ``Estrategia de  sostenibilidad 2014-2015, Universidad de
 Antioquia''. The authors also thank Tom\'as Verdugo, Stefan
 Gottloeber and Nelson Padilla for their comments on earlier version
 of this work.

\bibliographystyle{apj}
%\bibliography{references}
\begin{thebibliography}{15}
\expandafter\ifx\csname natexlab\endcsname\relax\def\natexlab#1{#1}\fi

@@

>>>>

The concentration of the NFW profile provides a conceptual framework
to study simulated dark matter halos as a function of redshift and
cosmological parameters.  In many numerical studies
\citep{Neto2007,Maccio2008,Duffy2008,Munoz2011,Prada2012,Ludlow2014,Ludlow2016,Klypin2016}
the results are summarized through the mass-concentration
relationship; that is the distribution of concentration values at a
fixed halo mass and redshift.  The success of such numerical
experiments rests on a reliable algorithm to estimate the
concentration.  Such an algorithm should provide unbiased results and
must be robust when applied to simulations of different numerical
resolution.

@@

To date, there are two standard methods to estimate concentrations in
dark matter halos extracted from N-body simulations.  The first method
tries to directly estimate the density profile.  It takes all the
particles in the halo and bins them in the logarithm of the radial
coordinate from the halo center.  Then, it estimates the density in
each logarithmic bin counting the particles and dividing by the
corresponding shell volume.  At this point is possible to make a
direct fit to the density as a function of the radial coordinate.
This method has been broadly used for more than two decades to study
the mass-concentration-redshift relation of dark matter halos. A
recent example is \citep{Ludlow2016}

A second method uses the circular velocity profile.  It finds the
value of $x$ for which the normalized circular velocity $v(<x)$ shows
a maximum.  Using this value it solves numerically for the
corresponding value of the concentration using Eq. (\ref{eq:max_v}).
A recent example of this methodology in cosmological simulation is
given by \cite{Klypin2016}.

@@

In our bootstrapping experiments we take halos with at least
$10^{5}$ particles and subsample them by factors of $2$ up to
$10^{3}$ to measure the concentration in every resampling.
We use a two-sample Kolmogorov-Smirnov (KS) test to check that the
bootstrapped sets are a fair subsample of its parent.
Using the KS test we compare the list of radial distances from each
subsample against that of its parent halo.
We find that the resulting p-value distribution is flat.
This confirms that the radial distribution of particles in the
bootraped halo is consistent with coming from the distribution given
by the parent halo.

@@

How does this impact the most recent mass-concentration estimates? 
\cite{Ludlow2016} and \cite{Klypin2016} estimated the mass-concentration
relation over a suite of cosmological N-body simulations using the
density and velocity methods, respectively. 
Both used halos that had at least $5\times 10^{3}$ particles,
so their results should not be affected by the bias we report here.
However, Figure 16 in \cite{Klypin2016} seems to push the results
below this particle limit for redshifts larger than $z=2.5$ where
their simulations should not be able to reach the mass range
$5\times 10^{10}$\hMsun - $5\times 10^{12}$\hMsun that they claim. 
Consistent with this limitation \cite{Ludlow2016} only report results
up to a redshift of $z=3$.

@@

 The authors acknowledge the technical support from the new
 high-performance computing facility at Uniandes. JEFR acknowledges
 financial support from Vicerrector\'ia de Investigaciones at Uniandes
 through a FAPA project. JCMC acknowledges financial support from
 ``Estrategia de  sostenibilidad 2014-2015, Universidad de
 Antioquia''. The authors also thank Tom\'as Verdugo, Stefan
 Gottloeber and Nelson Padilla for their comments on earlier versions
 of this work.

\bibliographystyle{apj}
%\bibliography{references}
\begin{thebibliography}{16}
\expandafter\ifx\csname natexlab\endcsname\relax\def\natexlab#1{#1}\fi

@@

\bibitem[{{Ludlow} {et~al.}(2016){Ludlow}, {Bose}, {Angulo}, {Wang},
  {Hellwing}, {Navarro}, {Cole}, \& {Frenk}}]{Ludlow2016}
{Ludlow}, A.~D., {Bose}, S., {Angulo}, R.~E., {Wang}, L., {Hellwing}, W.~A.,
  {Navarro}, J.~F., {Cole}, S., \& {Frenk}, C.~S. 2016, \mnras, 460, 1214
